Tangut translation is an extreme low-resource problem: in the current repository snapshot, only 491 real Tangut-Chinese pairs are available, with 400/41/50 train-dev-test examples. The task is also structurally unusual because many targets are short, title-like Chinese strings rather than ordinary modern sentences. We therefore evaluate four method families in the same setting: local prompt-only baselines, a stronger proprietary frontier prompt-only baseline, synthetic-data supervised fine-tuning (SFT), and direct preference optimization (DPO). The new frontier baseline, DeepSeek-V3.2 with dictionary grounding and few-shot title-oriented prompting, is the strongest single non-reference system under our reference-aware judge, reaching 2.80/5 overall with 12/50 exact matches. A paper built around ``stronger CoT still fails'' would therefore be wrong. The positive local result instead lies inside the trainable regime: structured multitask SFT is the strongest pure supervised baseline, and stricter gap-filtered DPO on top of that base yields the strongest local model. With only 498 preference pairs whose reward gap is at least 0.4, the sigmoid variant reaches chrF++ 30.01, keeps 5/50 exact matches, preserves 22/27 title suffixes, and stays near reference length. More importantly, a best-of-two oracle between the frontier baseline and this local model reaches 13/50 exact matches, 40.16 chrF++, and 3.06 reference-aware overall, showing that local adaptation contributes complementary hypotheses rather than merely weaker copies of the frontier output. Paired bootstrap resampling confirms that the local chrF++ gain over multitask SFT is robust (+4.02, 95\% CI [0.84, 7.80]), while the frontier model retains a robust advantage over the best local model on reference-aware overall (-0.58 when measured as local minus frontier, 95\% CI [-1.04, -0.12]) and exact match (-0.14, 95\% CI [-0.26, -0.04]). We also show that several apparently reasonable reference-free metrics are misaligned with the task: the gold human reference itself receives only moderate dictionary coverage and extremely poor perplexity under a modern-Chinese language model, despite achieving perfect chrF++ and perfect reference-aware judge scores. Finally, an overlap audit finds zero exact source, target, or pair overlap across train/dev/test splits and zero exact overlap between test references and the 50k synthetic multitask targets. These findings suggest a more precise practical lesson for ultra-low-resource Tangut short-text translation: frontier prompting is a strong proprietary upper bound, structured local supervision remains scientifically useful because it changes the error regime under severe data scarcity, and evaluation must be anchored to reference-aware metrics rather than generic fluency proxies.
