\section{Analysis}

\subsection{Reference-aligned diagnostics}

Repository-level metrics alone do not fully explain the new ranking. Additional diagnostics from stored predictions make the trade-off much clearer. The first important fact is that the frontier DeepSeek baseline is genuinely clean: it reaches 12/50 exact matches, no contamination, 24/27 suffix preservation, and a near-reference length ratio of 0.95. In other words, the frontier result is not just a high-capacity version of the verbose dictionary baseline. It is a strong short-title reconstruction system in its own right.

Within the local learned family, the multitask model remains the cleanest pure SFT system, with 5 exact matches, only 1 contaminated output, and a near-reference output length ratio (1.10). The looser gap-$0.2$ sigmoid DPO follow-up keeps the same 5 exact matches and drops contamination to 0/50, but stretches titles to 1.74 times the reference length. The looser robust+weighted follow-up instead preserves title suffixes most aggressively (23/27), but pays for that with 7 contaminated outputs.

The stricter gap-$0.4$ follow-ups change the surface behavior substantially. The strict sigmoid model keeps 5 exact matches, raises title-suffix preservation to 22/27, and restores the length ratio to 1.12 while reaching the highest local chrF++ in the repository (30.01). The strict robust+weighted model also keeps 5 exact matches and 22/27 suffix preservation, but with only 2 contaminated outputs instead of 7. In other words, stricter pair filtering does not merely preserve the earlier DPO gains; it regularizes them into something much closer to the short-title target distribution.

By contrast, the older local dictionary-prompting baseline often produces outputs that are too long and too explanatory. Its average output is about 1.83 times the reference length. This helps lexical coverage but hurts exact title reconstruction. The comparison between \texttt{baseline2} and the frontier DeepSeek row is therefore methodologically useful: prompt-only inference is not doomed here, but it requires a much stronger model and a much stricter title-oriented prompt than the local baseline setup.

\input{tables/diagnostics}

\subsection{Oracle complementarity and workflow value}

The strongest new result is not another single-system win, but the size of the best-of-two upper bound. Table~\ref{tab:oracle-complementarity} compares the strongest frontier system against the strongest local system and then reports separate item-level oracles for exact match, chrF++, and the reference-aware overall score.

\input{tables/oracle_complementarity}

The oracle numbers are much larger than either system alone. The exact-match union rises from 12/50 for the frontier baseline and 5/50 for the best local model to 13/50 overall. More importantly, the corpus chrF++ oracle rises to 40.16 and the reference-aware overall oracle rises to 3.06. This gap is too large to dismiss as noise or a trivial restatement of the pairwise win/loss counts. It means the local model is not merely a worse copy of the frontier model; it is generating hypotheses that are often different and sometimes decisively useful. Indeed, the two systems produce identical outputs on only 4/50 items, while on the remaining 46 items the better of the two reaches mean judge overall 2.89.

This changes the practical reading of the paper. The strongest deployment claim is no longer ``use the local model instead of a frontier API.'' It is instead ``if a frontier API is available, the local model is still useful as a complementary hypothesis generator.'' That is a stronger scholar-facing story than simple replacement: one can generate a frontier candidate, generate a local candidate, and then either ask a judge to arbitrate or surface the disagreement to a human annotator.

\subsection{A direct test of stronger CoT prompting}

The obvious objection to this project is straightforward: why not simply use a stronger Chinese frontier LLM with better chain-of-thought prompting? The new DeepSeek-V3.2 few-shot baseline lets us answer that objection directly. The short answer is that strong frontier prompting is indeed very effective. On the reference-aware judge, it reaches overall 2.80, above the best local score of 2.22, and in direct pairwise comparison it beats the strict gap-$0.4$ sigmoid model on 20 items, ties on 23, and loses on 7.

But this does not make the local-model story empty. The frontier advantage is concentrated on canonical or bibliographically regular titles, where broad Chinese title priors help a great deal. It is exact or near-exact on cases such as ``\zh{百千印陀羅尼經}'', ``\zh{同音}'', ``\zh{類林}'', and ``\zh{地藏菩薩本願經}'', where the best local model drifts badly. At the same time, the local model still wins a non-trivial minority of hard cases that are short, elliptical, or lexically awkward: it preserves ``\zh{五部經}'' where the frontier model normalizes it to ``\zh{五類經}''; it stays close to ``\zh{正理滴之句義顯具}'' while the frontier model rewrites it as ``\zh{德宜點義句顯宣論}''; it translates ``\zh{到賢}'' as ``\zh{達賢}'' while the frontier model hallucinates ``\zh{五台山觀音普賢殿寺}''; and it exactly recovers ``\zh{番言金剛王乘根}'' while the frontier model collapses to ``\zh{番}''.

The comparison therefore sharpens the paper rather than invalidating it. A paper claiming that ``stronger CoT still cannot do the task'' would now be wrong. A stronger and more honest claim is that frontier prompting is a proprietary upper bound, while structured local adaptation remains scientifically useful because it changes the error regime, remains trainable and reproducible, and supplies complementary hypotheses that substantially enlarge the best-of-two upper bound.

\subsection{Reference-aware judging changes the ranking}

The legacy LLM judge in the repository is reference-free, so we reran evaluation with a stricter judge that sees the Tangut input, the gold reference, and the candidate output simultaneously. This rerun serves two purposes. First, it checks whether the human reference is correctly ranked when the judge has the information it should need. Second, it tests whether the raw-metric ranking survives once the judge can penalize explanatory overgeneration and content drift directly.

\input{tables/reference_aware_eval}

Table~\ref{tab:reference-aware-eval} changes the paper narrative once more. The human reference still receives perfect 5.0 scores on all dimensions, so the judge is well anchored. Among non-reference systems, the frontier DeepSeek baseline is now the clear winner at 2.80 overall. Within the local learned family, however, the best overall score is shared by three multitask-base DPO variants: the loose sigmoid run, the strict sigmoid run, and the strict robust+weighted run all reach 2.22. Within that local tie, the strict gap-$0.4$ runs are stronger on title reconstruction: the strict sigmoid model reaches the highest local reference-agreement score (2.12) and the highest local source-faithfulness score (2.26), while the strict robust+weighted model reaches the highest local title-style score (4.70). The older \texttt{final\_v2} checkpoint remains worse than \texttt{baseline3\_2\_multitask}, so the positive DPO result is specific to the stronger base and higher-quality pairs, not to DPO in the abstract.

The subset breakdown makes the new trade-off sharper. On the 27 reference titles containing common title suffixes, the frontier model reaches mean overall score 3.15, well above the best local runs (2.41). On the 23 non-suffix items, however, the gap narrows: the frontier model reaches 2.39, while the looser \texttt{final\_gap02\_multitask\_sigmoid} reaches 2.30. Within the local family, the strict gap-$0.4$ variants dominate the suffix-bearing subset, clearly above \texttt{baseline3\_2\_multitask} (2.15), and they do so without the earlier surface penalties: the strict sigmoid run stays near reference length (1.12) and the strict robust run drives contamination back to 0.0\% on this subset. On the non-suffix items, the looser gap-$0.2$ sigmoid variant remains strongest among local models. This is strong evidence that prompt-only reasoning and local parameter adaptation are exploiting different kinds of regularity.

\subsection{Human-reference anchor}

One of the strongest pieces of evidence in the repository is the \texttt{human\_reference} experiment. The gold reference receives perfect chrF++ and perfect reference-aware judge scores, as expected, but only moderate lexical coverage and catastrophic perplexity. This reveals a core evaluation issue: the current lexical coverage metric measures overlap with dictionary-derived semantics rather than agreement with the gold title, while the perplexity model penalizes compact title style and historical transliterations.

The implication is methodological, not cosmetic. If we were to select models purely by lexical coverage or perplexity, we would risk preferring systems that paraphrase, overgenerate, or stop poorly instead of systems that best reconstruct the true title. The new frontier baseline makes this even clearer: it is strong overall under the reference-aware judge, but not because it wins lexical coverage or perplexity outright. Conversely, the strict gap-$0.4$ sigmoid model reaches the best non-reference chrF++ while remaining far below the frontier model on the reference-aware judge. Therefore, in the paper, lexical coverage and perplexity should be reported as supporting diagnostics, not as the primary decision criteria.

\subsection{Statistical robustness and data hygiene}

\input{tables/statistical_robustness}

The 50-example test set is too small to justify raw point estimates without uncertainty. The bootstrap results in Table~\ref{tab:statistical-robustness} clarify which claims are stable. The local chrF++ gain from multitask SFT to strict gap-$0.4$ sigmoid DPO survives resampling, with a 95\% confidence interval entirely above zero. The frontier model's advantage over the best local model on reference-aware overall and exact match also survives resampling. By contrast, the chrF++ difference between those two systems does not: its 95\% interval crosses zero widely. This is exactly the kind of mixed picture that the paper should surface rather than hide. The correct interpretation is therefore not ``local beats frontier'' or ``frontier dominates local'' in the abstract, but ``different metrics support different deployment decisions, while the oracle shows the systems are complementary.''

The overlap audit addresses a separate reviewer concern. There is zero exact source, target, or source-target-pair overlap across the train/dev/test splits; zero exact overlap between test references and the 50k synthetic multitask targets; and zero exact overlap between the four frontier few-shot exemplars and the test references. Real-split character $n$-gram overlap is non-zero, which is expected in a domain where many titles share Buddhist morphemes and suffixes, but the synthetic multitask set is not memorizing the held-out titles: its mean maximum Jaccard overlap with test references is only 0.027 for character bigrams, 0.004 for trigrams, and 0.000 for 4-grams. That does not prove complete immunity from leakage in a broader sense, but it does rule out the simplest exact-copy explanation for the current results.

\subsection{Legacy DPO fails for different reasons than multitask-base DPO}

The old negative result is not a random bad run. The preference data itself is imperfect. The repository contains 3,867 preference pairs, including a small number of duplicate or invalid entries. More importantly, a quick audit against the original synthetic targets shows that the reward-chosen candidate is closer to gold only about 71\% of the time under a simple similarity proxy, and that this quality rises monotonically with reward gap: 62.3\% for gaps in $[0.05, 0.10]$, 66.3\% for $(0.10, 0.20]$, 75.4\% for $(0.20, 0.40]$, and 81.5\% for gaps $\ge 0.40$. In other words, the high-gap filter is not a cosmetic hyperparameter. It is directly isolating cleaner preference labels.

The legacy selector also excludes the strongest multitask model from the DPO-base search. That turns out to matter. Once we switch to the multitask base and keep only pairs with reward gap $\ge 0.2$, DPO stops being a blanket negative result and begins to beat the multitask SFT baseline under the reference-aware judge. Tightening the filter further to gap $\ge 0.4$ makes the mechanism much clearer. Relative to the loose gap-$0.2$ sigmoid model, the strict sigmoid model raises chrF++ from 20.93 to 30.01 and shrinks the length ratio from 1.74 to 1.12 while preserving the same 2.22 overall judge score. Relative to the loose robust+weighted model, the strict robust+weighted model reduces contamination from 7/50 to 2/50 and keeps the overall judge score at 2.22. Even a deterministic short-title normalization pass raises \texttt{final\_gap02\_multitask\_robustwpo} only to chrF++ 29.06, still below the raw strict-sigmoid result. The reward is therefore not useless, but low-gap pairs inject exactly the kind of surface noise that a short-title system can least afford.

\input{tables/dpo_audit}
