\section{Conclusion}

Tangut short-text translation is difficult here because the data are tiny and the desired outputs are compact historical titles that generic MT objectives, metrics, and rerankers model poorly. After strong dictionary-grounded frontier prompting is already in place, local training remains useful only when it supplies conservative complementary titles that open rerankers can exploit. The new controls sharpen that lesson: simple vote already helps, guarded selection is the strongest clean open workflow, and generic pairwise MBR overshoots. The supplementary oracle-bone result now reinforces the same workflow-level point in a second script. On a small canonicalized OBIMD$\rightarrow$EVOBC stress test, dictionary-grounded prompting remains the best single model, but a simple open 2-way hybrid with MT-SFT still lifts exact match from 40/200 to 44/200 and reduces contamination to 0.5\%. The title-shaped workflow components should nevertheless still be treated as Tangut-specific until they are validated on richer oracle references. This paper stops at controlled comparison rather than claiming scholar-facing deployment readiness. Ultra-low-resource historical translation is better framed here as workflow design under reference-aware evaluation than as a single-model race or a premature usability study. The immediate next step is to distill the closed adjudication signal into stronger open reranking, expand the oracle side from canonicalized probes to matched oracle short-text references, and only then run expert-facing validation as a follow-up external check on the narrowed workflow.
