\section{Discussion and Limitations}

The practical reading of the paper is simple. If a user can make only one model call, the DeepSeek-V3.2 frontier baseline is the best single model we measured. But if the target use case is scholar-facing short-title reconstruction rather than single-shot API convenience, the stronger claim is the workflow claim: a catalog-aware adjudicator over frontier and local candidates is better than frontier alone.

That distinction makes the contribution more general, not less. The local models matter not because they beat a frontier model head-to-head, but because they diversify the candidate pool precisely where fluent modern normalization is risky. This recipe should transfer beyond Tangut book titles: train structurally conservative local models, use them as hypothesis generators, and adjudicate with a title-aware reranker rather than a generic fluency prior.

The scope should also be explicit. The current dataset supports Tangut short-text translation, especially title-like strings, not general open-domain Tangut sentence translation. The strongest hybrid gains are concentrated on short and mid-length items, while long titles remain hard for every system. This is also why we do not foreground classical multilingual-transfer baselines: such methods are strongest when pretrained models already cover the source script or nearby related languages, and the present repository offers neither in a reproducible form. Under those conditions, lexicon-grounded synthesis and structured local adaptation are the more defensible comparison family. The same caveat applies to the synthetic corpus: it is a structural training proxy, not realistic pseudo-Tangut. The new deterministic open reranker should likewise be read as a first reproducible hybrid baseline rather than the final word on open adjudication.

Two limitations matter most. First, the real dataset is extremely small, so the paper cannot claim stable behavior beyond this narrow domain. Second, both the adjudicator and the primary reference-aware judge currently rely on proprietary frontier models, and we do not yet provide a blinded human validation study for the judge. The local Qwen judge and deterministic open reranker reduce but do not remove this concern. To narrow the reproducibility gap, the repository should release the exact frontier prompting script, fixed few-shot exemplar IDs, adjudicator templates, per-item predictions, adjudicator outputs, and pair-filtering code. The most useful next experiments are therefore not ``more DPO'' in the abstract, but distilling the adjudicator into an open local reranker, validating the judge on a small expert-annotated subset, adding title-aware decoding controls, and testing transfer to other historical short-form translation settings.
