\section{Introduction}

Historical translation at the ultra-low-resource end is difficult for a different reason than ordinary low-resource MT. The challenge is not only that parallel data are scarce, but also that the desired outputs are often compact titles, glosses, or catalog strings rather than fluent modern sentences. A system can therefore sound plausible while still failing at the real scholarly task, which is to recover the right historical title in the right form. Tangut is a sharp case. In the repository studied here, only 491 real Tangut-Chinese pairs are available, and many targets are short title-like strings rather than ordinary modern sentences \cite{sommerschield2023ancient, wang2023evahan, domingo2018modernizing}.

These coupled constraints create two concrete risks. On the modeling side, local training can easily drift toward synthetic artifacts or explanatory paraphrase instead of compact title reconstruction. On the evaluation side, generic MT metrics can reward plausible semantic overlap while still missing the historical title itself.

This changes the paper's central question. Once a strong frontier Chinese LLM is given dictionary glosses and a title-oriented prompt, the interesting issue is no longer whether a local model can simply beat it head-to-head. The more useful question is what remains worth learning locally, and whether local hypotheses can improve a scholar-facing workflow even when frontier prompting is already strong.

We therefore organize the paper around three research questions:
\begin{enumerate}
\item \textbf{RQ1}: How far can strong frontier prompting go as a single-model upper bound?
\item \textbf{RQ2}: If local training still matters, what kind of supervision signal helps most?
\item \textbf{RQ3}: Do frontier and local systems fail differently enough to yield a reproducible workflow gain?
\end{enumerate}

We answer these questions by comparing local dictionary-grounded prompting, a deliberately strong frontier prompt-only baseline, synthetic pseudo-parallel SFT, preference optimization on top of SFT, and inference-time selection over frontier and local candidates \cite{chen2017incorporating, sennrich2016monolingual, rafailov2023dpo}. The experimental design is intentionally layered: first establish the strongest single-model upper bound, then identify the best open/local recipe, and only then ask whether complementarity can be turned into a workflow gain. Because the test set contains only 50 items and the numerically strongest adjudicator is closed, we state that final claim conservatively: the deterministic open selector is the main reproducible result, while the closed adjudicator is used as an upper bound on recoverable complementarity rather than as the paper's central deployable artifact.

The story that emerges is simple. Strong frontier prompting already sets a high single-model bar, so the paper cannot rely on a weak-baseline narrative. Local training still matters, but only when it is disciplined enough to preserve title structure: structured multitask supervision is the right base, and DPO helps only after both the base model and the preference pairs are cleaned aggressively. The main positive result is therefore not local replacement of the frontier system, but recovery of complementary conservative hypotheses through open selection, with closed adjudication used only to estimate remaining headroom. At the same time, lexical coverage and perplexity mis-rank the gold reference itself, so generic reference-free signals are unsafe as primary evidence.

The paper makes three contributions:
\begin{enumerate}
\item a research-question-driven case study that separates three claims often blurred together in low-resource work: the strongest single-model upper bound, the best open/local trainable recipe, and the extent to which complementary local hypotheses can be recovered in a workflow under only 491 real pairs;
\item a positive local-method result showing that structured multitask supervision is the right base, that DPO helps only after aggressive preference-gap filtering, and that local models matter because they contribute complementary conservative hypotheses rather than because they dominate the frontier model head-to-head;
\item a reference-aware evaluation and label-quality analysis showing why short historical titles must not be judged primarily by generic reference-free metrics, why low-gap preference data destabilize alignment, and how much complementarity an open selector can recover reproducibly.
\end{enumerate}

The rest of the paper follows the research questions in that order and keeps two final claims separate on purpose: how to build the strongest open Tangut model, and how much scholar-facing benefit can be recovered from frontier/local complementarity without overclaiming beyond the evidence.
