\section{Methods}

\subsection{Inference-only baselines}

The first family contains no weight updates. \texttt{baseline1} prompts the local model directly on the Tangut string; \texttt{baseline2} adds dictionary glosses; and \texttt{baseline2\_1\_cot} further decomposes generation into alignment, literal drafting, and rewriting. This family is closest to lexicon-augmented MT, except that the lexicon is injected through prompting rather than decoder constraints \cite{chen2017incorporating, hokamp2017lexically}.

We also add a deliberately stronger prompt-only comparison based on DeepSeek-V3.2 with dictionary grounding and four fixed development-set exemplars (IDs 1, 6, 15, and 16: \zh{佛説大人八覺經}, \zh{佛説菩薩行修經}, \zh{金光明總持經}, and \zh{摩訶般若波羅蜜多心經}). These cover a short canonical sutra title, a bodhisattva-practice title, a dharani-style title, and a long transliteration-heavy chain. The prompt emphasizes compact title form, forbids explanatory paraphrase, and emits only the final title string. Decoding uses temperature 0.0 with a 512-token cap. The goal is not to build a straw baseline but to test directly whether frontier prompting already solves the task well enough.

\subsection{Synthetic SFT baselines}

The second family uses synthetic training data. We build 50k pseudo-Tangut/Chinese pairs from monolingual historical Chinese by mapping Chinese characters to Tangut characters through dictionary explanations, fixing the synthesis seed to 42, and keeping source lengths between 4 and 64 characters. The four branches differ only in how dictionary misses are handled and whether the target includes alignment structure: mixed-script keeps 30--70\% Chinese characters, \texttt{UNK} replaces misses with \texttt{[UNK]}, semantic projection replaces misses through Qwen2.5-0.5B embeddings plus FAISS lookup, and multitask keeps the \texttt{UNK} input but changes the target to \texttt{<dict\_match>\{...\}</dict\_match> <literal>\{...\}</literal> \{final title\}}. This gives the model both an alignment channel and a final title target in one sequence. These examples are not philologically faithful Tangut reconstructions; they are structured training proxies. Each synthetic set is combined with the 400 real training examples upsampled by a factor of 10 and shuffled with seed 42. The key design question is whether the model benefits more from input purity, approximate semantic preservation, or explicit alignment structure.

\subsection{Preference optimization}

The final family applies DPO to an SFT base using reward-defined chosen/rejected pairs \cite{rafailov2023dpo}. We sample five outputs per training prompt at temperature 0.8 and top-$p$ 0.95, then score them with
\[
r(y) = \mathrm{lex}(y) - 0.01 \log \mathrm{ppl}(y),
\]
where $\mathrm{lex}(y)$ is the repository's lexical-coverage score. Concretely, lexical coverage segments the Chinese output with jieba plus single-character backoff, performs maximum forward matching over the Tangut input with dictionary entries up to length five, and reports the fraction of matched Tangut spans whose gloss set intersects the output token set. The $\log \mathrm{ppl}(y)$ term is down-weighted by 0.01 so that fluency remains secondary. We then keep the top-scoring output as chosen and the bottom-scoring output as rejected whenever the reward gap is at least 0.05.

Because reviewers reasonably questioned the perplexity term, we also ablate the coefficient in $\mathrm{lex}(y) - w \log \mathrm{ppl}(y)$ for $w \in \{0, 0.005, 0.01, 0.02\}$. This produces zero pair-order flips relative to $w = 0.01$ and changes the retained high-gap pool ($\ge 0.4$) only from 498 to 504 pairs, so perplexity acts only as a weak tie-breaker in the cleaned data rather than the main preference signal.

The legacy pipeline first selects the best SFT base among the mixed, \texttt{UNK}, and semantic variants using chrF++, then constructs preference pairs from that base. In the current snapshot that selector chooses the \texttt{UNK} model, yielding the legacy checkpoint \texttt{final\_v2}, which serves as a negative control. The follow-up DPO runs instead start from the stronger multitask SFT model. After auditing the preference data, we keep only pairs whose reward gap exceeds either 0.2 (1,996 pairs) or 0.4 (498 pairs), and we train both standard sigmoid DPO and a robust+weighted variant \cite{chen2024robustdpo,pang2024wpo}. This turns the DPO family into a controlled test of base-model choice, pair filtering, and loss robustness.

\subsection{Catalog-aware hybrid adjudication}

The oracle analysis suggested that frontier and local candidates are complementary, but an oracle is not a deployable system. We therefore build a real inference-time adjudicator over the frontier DeepSeek baseline, UNK-SFT, and MT-DPO g0.4 sig. For each item it sees the Tangut source, dictionary glosses, the three candidate titles, their lengths, and simple prefix/suffix consensus hints.

The adjudicator uses Azure-hosted GPT-5.4 at temperature 0.0. Its prompt treats the task as historical catalog-title reconstruction rather than free paraphrase: dictionary glosses are clues rather than canonical modern titles, and short inputs are penalized for explanatory expansion. The model may either copy one candidate or perform a minimal repair using only candidate fragments or directly attested gloss segments. The repository stores the exact templates, frontier prompting script, fixed exemplar IDs, and per-example adjudication JSON, so the closed-model part of the workflow is at least auditable.

For reproducibility, we also add a deterministic open selector over the same 3-way pool. It never synthesizes text or calls an external model. It scores candidates using pairwise prefix/suffix consensus (shared prefix or suffix lengths counted only when they are at least two characters long), a title-suffix prior for final characters in \{\zh{經論記疏頌儀義傳錄贊序字品文觀根次門}\}, deviation from the candidate-median length $m$, contamination penalties for ASCII spillover, angle brackets, \texttt{[UNK]}, or known prompt artifacts, anti-truncation ($|y| \le \max(1, |x|/4)$) and anti-expansion ($|y| \ge m + 5$ and $|y| \ge |x| + 5$) guards, and a small frontier prior. A guarded switch replaces the frontier output only when a local candidate has materially stronger conservative evidence. Table~\ref{tab:oracle-complementarity} later tests a 4-way extension that adds MT-SFT and shows that the original 3-way pool is already the best conservative operating point.

\subsection{Training and judge details}

All learned models start from Qwen2.5-7B-Instruct with LoRA adapters. SFT uses three epochs, learning rate $2\times10^{-4}$, batch size 4, gradient accumulation 4, rank-64 LoRA, and maximum sequence length 512. DPO uses two epochs, learning rate $5\times10^{-5}$, batch size 1, gradient accumulation 8, $\beta=0.1$, rank-32 LoRA, cosine decay, and the same sequence length; the robust variants add label smoothing 0.1 and WPO-style weighting. The full pipeline fixes seed 42 for synthesis, data combination, and DPO training.

The reference-aware judge is a supplementary evaluator, not a training signal. It uses Azure-hosted GPT-5.4 to score reference agreement, source faithfulness, title-style fitness, and overall quality on a 1--5 scale after seeing the Tangut source, gold reference, and candidate simultaneously. Final scores are simple means over examples. For cross-judge robustness, the repository also includes a local Qwen2.5-7B-Instruct judge with the same rubric. We use the Azure judge as the primary table because it is the stricter and less saturated scorer.
