\begin{thebibliography}{10}
\small

\bibitem{sommerschield2023ancient}
Thea Sommerschield et al.
\newblock 2023.
\newblock Machine Learning for Ancient Languages: A Survey.
\newblock \emph{Computational Linguistics}, 49(3).

\bibitem{wang2023evahanoverview}
Wang et al.
\newblock 2023.
\newblock EvaHan2023: Overview of the First International Ancient Chinese Translation Bakeoff.
\newblock In \emph{Proceedings of ALT2023: Ancient Language Translation Workshop}.

\bibitem{domingo2018backtranslation}
Miguel Domingo and Francisco Casacuberta.
\newblock 2018.
\newblock A Machine Translation Approach for Modernizing Historical Documents Using Backtranslation.
\newblock In \emph{Proceedings of IWSLT}.

\bibitem{sennrich2016monolingual}
Rico Sennrich, Barry Haddow, and Alexandra Birch.
\newblock 2016.
\newblock Improving Neural Machine Translation Models with Monolingual Data.
\newblock In \emph{Proceedings of ACL}.

\bibitem{arthur2016lexicons}
Philip Arthur, Graham Neubig, and Satoshi Nakamura.
\newblock 2016.
\newblock Incorporating Discrete Translation Lexicons into Neural Machine Translation.
\newblock In \emph{Proceedings of EMNLP}.

\bibitem{gu2018universal}
Jiatao Gu et al.
\newblock 2018.
\newblock Universal Neural Machine Translation for Extremely Low Resource Languages.
\newblock In \emph{Proceedings of NAACL-HLT}.

\bibitem{hokamp2017lexically}
Chris Hokamp and Qun Liu.
\newblock 2017.
\newblock Lexically Constrained Decoding for Sequence Generation Using Grid Beam Search.
\newblock In \emph{Proceedings of ACL}.

\bibitem{ranathunga2023survey}
Surangika Ranathunga et al.
\newblock 2023.
\newblock Neural Machine Translation for Low-Resource Languages: A Survey.
\newblock \emph{ACM Computing Surveys}, 55(11).

\bibitem{lu2024cod}
Hongyuan Lu et al.
\newblock 2024.
\newblock Chain-of-Dictionary Prompting Elicits Translation in Large Language Models.
\newblock In \emph{Proceedings of EMNLP}.

\bibitem{liao2024lfm}
You-Cheng Liao et al.
\newblock 2024.
\newblock Learning-From-Mistakes Prompting for Indigenous Language Translation.
\newblock In \emph{Proceedings of the Seventh Workshop on Technologies for Machine Translation of Low-Resource Languages}.

\bibitem{rafailov2023dpo}
Rafael Rafailov et al.
\newblock 2023.
\newblock Direct Preference Optimization: Your Language Model is Secretly a Reward Model.
\newblock In \emph{Proceedings of NeurIPS}.

\bibitem{chowdhury2024robustdpo}
Sayak Ray Chowdhury, Anush Kini, and Nagarajan Natarajan.
\newblock 2024.
\newblock Provably Robust DPO: Aligning Language Models with Noisy Feedback.
\newblock In \emph{Proceedings of ICML}.

\bibitem{zhou2024wpo}
Wenxuan Zhou et al.
\newblock 2024.
\newblock WPO: Enhancing RLHF with Weighted Preference Optimization.
\newblock In \emph{Proceedings of EMNLP}.

\bibitem{uhlig2024dqo}
Tobias Uhlig et al.
\newblock 2024.
\newblock Cross-lingual Human-Preference Alignment for Neural Machine Translation with Direct Quality Optimization.
\newblock arXiv preprint arXiv:2409.17673.

\bibitem{yang2024dpombr}
Yang et al.
\newblock 2024.
\newblock Direct Preference Optimization for Neural Machine Translation with Minimum Bayes Risk Decoding.
\newblock In \emph{Proceedings of NAACL}.

\bibitem{popovic2017chrfpp}
Maja Popovic.
\newblock 2017.
\newblock chrF++: Words Helping Character n-grams.
\newblock In \emph{Proceedings of the Second Conference on Machine Translation}.

\bibitem{rei2020comet}
Ricardo Rei et al.
\newblock 2020.
\newblock COMET: A Neural Framework for MT Evaluation.
\newblock In \emph{Proceedings of EMNLP}.

\bibitem{zouhar2024cometpitfalls}
Vilem Zouhar et al.
\newblock 2024.
\newblock Pitfalls and Outlooks in Using COMET.
\newblock In \emph{Proceedings of WMT}.

\bibitem{pombal2025mint}
Jose Pombal et al.
\newblock 2025.
\newblock Adding Chocolate to Mint: Mitigating Metric Interference in Machine Translation.
\newblock \emph{Transactions of the Association for Computational Linguistics}.

\bibitem{kepler2019openkiwi}
Fabio Kepler et al.
\newblock 2019.
\newblock OpenKiwi: An Open Source Framework for Quality Estimation.
\newblock In \emph{Proceedings of ACL: System Demonstrations}.

\end{thebibliography}
