\section{Related Work}

\paragraph{Ancient and historical language processing.}
Ancient-language NLP differs from mainstream NLP not only because data are scarce, but also because scripts, editorial conventions, and annotation practices are unstable across corpora \cite{sommerschield2023ancient}. Recent ancient-Chinese translation bakeoffs and historical-document modernization studies operate under similar sparsity, though they typically target sentence translation or orthographic updating rather than cross-script title reconstruction \cite{wang2023evahanoverview, domingo2018backtranslation}. Our Tangut setting is narrower still: only 491 real pairs are available, and the evaluation set is dominated by compact title-like strings.

\paragraph{Low-resource MT, synthetic supervision, and lexical grounding.}
Low-resource MT often relies on synthetic parallel data, multilingual transfer, lexicon integration, or explicit decoding constraints \cite{sennrich2016monolingual, gu2018universal, ranathunga2023survey, arthur2016lexicons, hokamp2017lexically}. Tangut overlaps with this literature but is harsher in two ways: the source script is largely outside standard MT pretraining, and the outputs are compact catalog titles rather than ordinary sentences. We therefore focus on synthetic supervision and lexicon-grounded structure rather than on broad multilingual transfer.

\paragraph{Dictionary-structured prompting and error-aware prompting.}
Recent LLM translation work suggests that lexical resources are most useful when presented as structured intermediate evidence rather than as flat prompt context. Chain-of-Dictionary prompting and Learning-From-Mistakes prompting both follow this pattern in low-resource settings \cite{lu2024cod, liao2024lfm}. Our stronger frontier baseline is adjacent to this line, but the present paper asks a different question: whether alignment structure should be absorbed into trainable local targets, and whether those local hypotheses remain useful after a strong frontier prompt-only baseline is already in place.

\paragraph{Preference optimization.}
Direct Preference Optimization (DPO) is now a standard offline alignment objective, and recent work has adapted direct preference or quality optimization to machine translation while also studying robustness under noisy feedback \cite{rafailov2023dpo, uhlig2024dqo, yang2024dpombr, chowdhury2024robustdpo, zhou2024wpo}. Our results place this literature in a harsher regime: supervision is tiny, outputs are title-like, and the chosen/rejected signal is built from partially misaligned automatic metrics. In this setting, DPO helps only after base-model choice and pair filtering are controlled aggressively.

\paragraph{Evaluation for non-standard translation outputs.}
chrF++ is widely used in machine translation because character-level overlap is often more stable than token-level overlap under orthographic variation \cite{popovic2017chrfpp}. Learned MT metrics and QE systems such as COMET and OpenKiwi go beyond surface overlap, but recent work also shows that they can have domain-specific failure modes and induce metric interference \cite{rei2020comet, kepler2019openkiwi, zouhar2024cometpitfalls, pombal2025mint}. Our findings sharpen that caution in a historical short-text regime: lexical coverage and perplexity can penalize the gold reference itself, so reference-aware evidence must remain primary.
