\section{Main Results}

Table~\ref{tab:main-results} reports only representative systems. The empirical story is short: a strong frontier baseline already answers much of the single-model question, local learning helps only when it stays close to title structure, and the main reproducible gain comes from selecting among complementary candidates rather than asking one local model to replace the frontier system.

\input{tables/main_results}

\subsection{RQ1: Is Strong Frontier Prompting Already Enough as a Single-Model Upper Bound?}

Yes. The DeepSeek frontier baseline is far stronger than the local prompt baselines, reaching 28.54 chrF++, 12/50 exact matches, and 2.80 overall under the reference-aware judge. Prompt-only is therefore not a straw baseline here. A strong frontier system is already hard to beat, so every later claim has to explain what local training still contributes after that bar is in place. The frontier configuration is also fully enumerated now: \texttt{deepseek/deepseek-v3.2} via OpenRouter, temperature 0.0, maximum 512 output tokens, and four fixed development-set exemplars. Because decoding is deterministic, sampling seeds are not relevant. We do not claim a full paid prompt-variance sweep, but the stored pilot variants in the repository support the chosen prompt direction: on the shared 5-item pilot slice, the final few-shot title-only prompt reaches 2/5 exact matches and 67.68 chrF++, versus 0/5 and 21.99 chrF++ for the strict title-only prompt.

\subsection{RQ2: If Local Training Still Matters, What Signal Helps Most?}

The answer is not that more local training is always better. What helps is supervision that explicitly teaches the model how a short historical title is assembled. UNK-SFT already improves over prompt-only local decoding by enforcing pure Tangut inputs, and multitask SFT improves again, making structured supervision the right base. DPO is not uniformly helpful: the legacy DPO checkpoint trained on the old UNK-selected base drops to 19.39 chrF++ and 1.68 overall, so reward-based alignment is clearly not a blanket win. Once DPO is moved onto the multitask base and filtered aggressively, however, it becomes useful: MT-DPO g0.4 sig is the strongest local model at 30.01 chrF++ and 2.22 overall. The gap-0.4 robustWPO control reaches the same overall score but lower chrF++, so the main gain is still better explained by cleaner labels and the stronger multitask base than by a more elaborate loss alone.

\subsection{RQ3: Can Complementary Failures Be Turned into a Stronger Workflow?}

Yes, and this is where the paper's main positive result appears. The frontier model remains the best single predictor, but the best local models are still useful because they fail differently and often more conservatively on short titles. Table~\ref{tab:oracle-complementarity} includes not only our guarded selector, but also standard open reranking controls over the same 3-way frontier/UNK/DPO pool. Even a trivial plurality vote already improves the frontier baseline to 30.06 chrF++, 13/50 exact matches, and 2.94 overall, which shows that complementarity is visible without any learned scorer. A local Qwen reranker reaches 30.85 chrF++ and 2.82 overall. Pairwise chrF-MBR pushes overlap higher to 34.76 chrF++, yet it over-switches and hurts judged quality, dropping to 2.64 overall. The guarded heuristic is therefore not the only useful open reranker, but it is the strongest clean overlap-oriented one, reaching 32.62 chrF++ and 2.90 overall with zero contamination. The 3-way catalog-aware adjudicator still goes further to 34.87 chrF++, 3.02 overall, and 14/50 exact matches, but we treat that result only as a closed upper bound on recoverable complementarity.

\input{tables/oracle_complementarity}

The workflow result should therefore be read in two layers. Operationally, the main reproducible claim is that open reranking over frontier and local candidates helps, not that one specific heuristic is uniquely best. Analytically, the closed adjudicator shows that the candidate pool contains additional complementarity that a stronger reranker can exploit. Pool ablations and sensitivity checks in the repository do not change this ordering.

\subsection{Supplementary Oracle Portability Probe}

Table~\ref{tab:oracle-portability} reports a deliberately conservative second-script stress test: a 200-item held-out canonicalized OBIMD$\rightarrow$EVOBC probe. The setup is narrower than the Tangut benchmark and intentionally under-retuned. It evaluates text-side decipherment over real OBIMD reading sequences after canonicalizing them into the EVOBC ID inventory through shared modern-character labels, and it does not reuse the Tangut-specific reranker or the reference-aware judge. Zero-shot Qwen largely refuses the surrogate input, collapsing to long disclaimers with 0/200 exact matches, 0.44 chrF++, and 27.5\% contamination. Once dictionary grounding is added, prompt-only transfer becomes immediately usable, reaching 40/200 exact matches and 46.20 chrF++ with only 3.5\% contamination.

\input{tables/oracle_portability_probe}

The key portability signal is therefore not local single-model dominance, but survival of complementary repair behavior under a harsher interface. The oracle multitask SFT checkpoint still emits a three-part output; if we score the raw assistant tail, it badly mismatches the compact oracle target, yielding 13.11 chrF++ with 100\% contamination. Reading the multitask \texttt{<literal>} field instead recovers the probe-aligned short-form output, reaching 29/200 exact matches and 40.56 chrF++. The dictionary-grounded prompt and MT-SFT \texttt{<literal>} outputs differ on 179/200 items, and a deliberately simple open 2-way selector already turns that complementarity into a reproducible gain, lifting exact match from 40/200 to 44/200, chrF++ from 46.20 to 47.81, and contamination from 3.5\% down to 0.5\%. This oracle result therefore supports, rather than competes with, the main Tangut story: what transfers most cleanly is not local single-model dominance, but the workflow claim that local training can still matter by supplying conservative alternative hypotheses once a strong prompt baseline is already in place.
